\frfilename{mt2a.tex} 
\versiondate{3.8.15} 
 
\def\RoverC{\hbox{\biggerscriptfonts${{\Bbb R}\atop{\Bbb C}}$}} 
 
\def\chaptername{Appendix} 
\def\sectionname{Introduction} 
 
\gdef\topparagraph{} 
\gdef\bottomparagraph{Appendix to Vol.\ 2 {\it intro.}} 
 
\centerline{\bf Appendix to Volume 2} 
 
\medskip 
 
\centerline{\bf Useful Facts} 
 
\medskip 
 
In the course of writing this volume, I have found that a considerable 
number of 
concepts and facts from various branches of mathematics are necessary to 
us.   Nearly all of them are embedded in important and well-established 
theories for which many excellent textbooks are available and which I 
very much hope that you will one day study in depth.   Nevertheless, I 
am reluctant to send you off immediately to courses in 
general topology, functional analysis and set theory, as if these were 
essential prerequisites for our work here, along with real analysis and 
basic linear algebra.   For this reason I have written this 
Appendix, setting out those results which we actually need at some point 
in this volume.   The great majority of them really are  
elementary -- indeed, some are so 
elementary that they are not always spelt out in detail in orthodox 
treatments of their subjects. 
 
While I do not put this book forward as the proper place to learn any of 
these topics, I have tried to set them out in a way that you will find 
easy to integrate into regular approaches.   I do not expect anybody to 
read systematically through this work, and I hope that the references 
given in the main chapters of this volume will be adequate to guide you 
to the particular items you need. 
 
\discrpage 
 
