\frfilename{mt43.tex}
\versiondate{6.11.04}
\copyrightdate{1997}

\def\chaptername{Topologies and measures II}
\def\sectionname{Introduction}

\newchapter{43}

The first chapter of this volume was `general' theory of topological
measure spaces;  I attempted to distinguish the most important
properties a topological measure can have -- inner regularity,
$\tau$-additivity -- and describe their interactions at an abstract
level.   I now turn to rather more specialized investigations, looking
for features which offer explanations of the behaviour of the most
important spaces, radiating outwards from Lebesgue measure.

In effect, this chapter consists of three
distinguishable parts and two appendices.   The first three sections are
based on ideas from descriptive set theory, in particular Souslin's
operation (\S431);  the properties of this operation are the foundation
for the theory of two classes of topological space of particular
importance in measure theory, the
K-analytic spaces (\S432) and the analytic spaces (\S433).
The second part of the chapter, \S\S434-435, collects
miscellaneous results on Borel and Baire measures, looking at the ways
in which topological properties of a space determine properties of the
measures it carries.   In \S436 I
present the most important theorems on the
representation of linear functionals by integrals;  if you like, this is
the inverse operation to the construction of integrals from measures in
\S122.   The ideas continue into \S437, where I discuss spaces of signed
measures representing the duals of spaces of continuous functions, and
topologies on spaces of measures.
The first appendix, \S438, looks at a
special topic:  the way in which
the patterns in \S\S434-435 are affected if we assume that our spaces
are not unreasonably complex in a rather
special sense defined in terms of measures on discrete spaces.
Finally, I end the chapter with a further collection of
examples, mostly to exhibit boundaries to the theorems of the chapter,
but also to show some of the variety of the structures we are dealing
with.

\discrpage

