\frfilename{mt498.tex}
\versiondate{4.9.08}
\copyrightdate{2002}

\def\chaptername{Further topics}
\def\sectionname{Cubes in product spaces}

\newsection{498\dvAformerly{\S4{}94}}

I offer a brief note on a special property of (Radon) product measures.

\leader{498A}{Proposition} Let $(\frak A,\bar\mu)$ be a semi-finite
measure algebra with its measure-algebra topology\cmmnt{ (323A)}.
Suppose that $A\subseteq\frak A$ is an uncountable analytic set.   Then
there is a compact set $L\subseteq A$, homeomorphic to
$\{0,1\}^{\Bbb N}$, such that $\inf L\ne 0$ in $\frak A$.

\proof{ $A\setminus\{0\}$ is still an uncountable analytic subset of
$\frak A$.   By 423J, it has a subset homeomorphic to
$\{0,1\}^{\Bbb N}\cong\Cal P\Bbb N$;  let
$f:\Cal P\Bbb N\to A\setminus\{0\}$ be an injective continuous
function.   Because $(\frak A,\bar\mu)$ is semi-finite, there is an $a\Bsubseteq f(\emptyset)$ such that $0<\bar\mu a<\infty$;  set $\delta=\bar\mu a$.   Note that
$(I,J)\mapsto\bar\mu(a\Bcap f(I)\Bsetminus f(J)):
(\Cal P\Bbb N)^2\to\Bbb R$ is continuous.   Choose a sequence
$\sequencen{k_n}$ in $\Bbb N$ inductively, as follows.
Given $\ofamily{i}{n}{k_i}$, set $K_n=\{k_i:i<n\}$.   For each
$J\subseteq K_n$ we have
$\lim_{r\to\infty}\bar\mu(a\Bcap f(J)\Bsetminus f(J\cup\{r\}))=0$, so there is a $k_n$, greater than $k_i$ for every $i<n$, such that $\bar\mu(a\Bcap f(J)\Bsetminus f(J\cup\{k_n\})\le 2^{-2n-2}\delta$ for every $J\subseteq K_n$;  continue.

Now

\Centerline{$\bar\mu(a\Bcap\inf_{J\subseteq K_n}f(J))
\ge\delta(\Bover12+2^{-n-1})$}

\noindent for every $n\in\Bbb N$.   \Prf\ Induce on $n$.   If $n=0$,
then $a\Bcap\inf_{J\subseteq K_n}f(J)=a\Bcap f(\emptyset)$ has measure
$\delta=\delta(\bover12+2^{-1})$.   For the inductive step to
$n+1\ge 1$, observe that

$$\eqalignno{\bar\mu(a\Bcap\inf_{J\subseteq K_{n+1}}f(J))
&=\bar\mu(a\Bcap\inf_{J\subseteq K_n}f(J)\Bcap f(J\cup\{k_n\}))\cr
&\ge\bar\mu(a\Bcap\inf_{J\subseteq K_n}f(J))
 -\sum_{J\subseteq K_n}
  \bar\mu(a\Bcap f(J)\Bsetminus f(J\cup\{k_n\}))\cr
&\ge\delta(\Bover12+2^{-n-1})
 -\sum_{J\subseteq K_n}2^{-2n-2}\delta\cr
\displaycause{by the inductive hypothesis and the choice of $k_n$}
&=\delta(\Bover12+2^{-n-2}).\cr}$$

\noindent So the induction proceeds.\ \Qed

Set $K=\{k_i:i\in\Bbb N\}$, $c=\inf\{f(J):J\subseteq K$ is finite$\}$.
Then

\Centerline{$\bar\nu(a\Bcap c)
=\inf_{n\in\Bbb N}\bar\mu(a\Bcap\inf_{J\subseteq K_n}f(J))
\ge\Bover12\delta$,}

\noindent and $c\ne 0$.   But now observe that
$L=f[\Cal PK]$ is a subset of $A$
homeomorphic to $\Cal PK$ and therefore to $\{0,1\}^{\Bbb N}$.
Also $\{b:b\Bsupseteq c\}$ is closed (323D(d-i)), so
$C=\{J:f(J)\Bsupseteq c\}$ is closed in $\Cal PK$;  as it
includes the dense set $[K]^{<\omega}$, $C=\Cal PK$ and
$\inf L\Bsupseteq c$ is non-zero.
}%end of proof of 498A

\leader{498B}{Proposition}\cmmnt{ (see {\smc Brodski\v{\i} 49},
{\smc Eggleston 54})}
Let $(X,\frak T,\Sigma,\mu)$ be an atomless Radon measure space,
$(Y,\frak S,\Tau,\nu)$ an effectively locally finite $\tau$-additive
topological measure space, and $\tilde\lambda$ the $\tau$-additive
product measure on
$X\times Y$\cmmnt{ (417G)}.   Then if $W\subseteq X\times Y$ and
$\tilde\lambda W>0$ there are a
non-scattered compact set $K\subseteq X$ and a set $F\subseteq Y$ of
positive measure such that $K\times F\subseteq W$.

\proof{{\bf (a)} To begin with (down to the end of (c)), let us suppose
that both $\mu$ and $\nu$ are totally finite.   Let $(\frak B,\bar\nu)$
be the measure algebra of $\nu$.   Because $\tilde\lambda$ is inner
regular with respect to the closed sets, we may suppose that $W$ is
closed.   In this case, writing $\lambda$ for the c.l.d.\ product
measure on $X\times Y$, there is a $W'\supseteq W$ such that
$\lambda W'$ is defined and equal to $\tilde\lambda W$
(417C(iv)).   By 418Tb, there is a $\mu$-conegligible set $X_0$ such
that $W'[\{x\}]\in\Tau$ for every $x\in X_0$,
$B=\{W'[\{x\}]^{\ssbullet}:x\in X_0\}$ is separable for the
measure-algebra topology of $\frak B$, and
$x\mapsto W'[\{x\}]^{\ssbullet}:X_0\to\frak B$ is measurable.   Now
Fubini's theorem, applied in the form of 252D to $\lambda$ and in the
form of 417Ha to $\tilde\lambda$, tells us that

\Centerline{$\int\nu W'[\{x\}]\mu(dx)=\lambda W'=\tilde\lambda W
=\int\nu W[\{x\}]\mu(dx)$.}

\noindent So
$X_1=\{x:x\in X_0$, $W'[\{x\}]^{\ssbullet}=W[\{x\}]^{\ssbullet}\}$ is
$\mu$-conegligible.
Since the topology of $\frak B$ is metrizable (323Ad or 323Gb),
$B$ is separable
and metrizable, and $x\mapsto W[\{x\}]^{\ssbullet}:X_1\to B$ is almost
continuous (418J, applied to the subspace measure on $X_1$).   Let
$K^*\subseteq X_1$ be a compact set of non-zero measure such that
$x\mapsto W[\{x\}]^{\ssbullet}:K^*\to\frak B$ is continuous.

\medskip

{\bf (b)} There is a non-zero $c\in\frak B$ such that
$K_c=\{x:x\in K^*$, $c\Bsubseteq W[\{x\}]^{\ssbullet}\}$ is compact and
not scattered.   \Prf\ Because $x\mapsto W[\{x\}]^{\ssbullet}$ is
continuous on $K^*$, $B^*=\{W[\{x\}]^{\ssbullet}:x\in K^*\}$ is compact
and every $K_c$ is compact.   (i) If $B^*$ is countable, then
$K^*=\bigcup_{b\in B^*}K_b$, so there is some
$c\in B^*$ such that $\mu K_c>0$.   Let $E$ be a non-negligible
self-supporting subset of $K_c$;  then (because $\mu$ is atomless,
therefore zero on singletons) $E$ has no isolated points.   So $K_c$ is
not scattered.   (ii) If $B^*$ is uncountable, then by
498A there is a set $D\subseteq B^*$, homeomorphic to
$\{0,1\}^{\Bbb N}$, with a non-zero lower bound $c$ in $\frak B$.   Now
$\{W[\{x\}]^{\ssbullet}:x\in K_c\}$ includes $D$, so $\{0,1\}^{\Bbb N}$
and therefore $[0,1]$ are continuous images of closed subsets of $K_c$
and $K_c$ is not scattered (4A2G(j-iv)).\ \Qed

\medskip

{\bf (c)} Set $K=K_c$.   Then 414Ac tells us that

\Centerline{$(\bigcap_{x\in K}W[\{x\}])^{\ssbullet}
=\inf_{x\in K}W[\{x\}]^{\ssbullet}\Bsupseteq c$}

\noindent is non-zero, so $F=\bigcap_{x\in K}W[\{x\}]$ is
non-negligible;  while $K\times F\subseteq W$.   So we have found
appropriate sets $K$ and $F$, at least when $\mu$ and $\nu$ are totally
finite.

\medskip

{\bf (d)} For the general case, we need observe only that by 417C(iii)
there are $X'\in\Sigma$ and $Y'\in\Tau$, both of finite measure, such
that $\tilde\lambda(W\cap(X'\times Y'))>0$.   Now the subspace measure
$\mu_{X'}$ on $X'$ is atomless and Radon (214Ka, 416Rb), the subspace
measure $\nu_{Y'}$ on $Y'$ is $\tau$-additive (414K), and the
$\tau$-additive product of $\mu_{X'}$ and $\nu_{Y'}$ is the subspace
measure on $X'\times Y'$ induced by $\tilde\lambda$ (417I).   So we can
apply (a)-(c) to $\mu_{X'}$ and $\nu_{Y'}$ to see that there are a
non-scattered compact set $K\subseteq X'$ and a non-negligible
measurable set $F\subseteq Y'$ such that $K\times F\subseteq W$.
}%end of proof of 498B

\leader{498C}{Proposition}\cmmnt{ (see
{\smc Ciesielski \& Pawlikowski 03})} Let
$\familyiI{(X_i,\frak T_i,\Sigma_i,\mu_i)}$ be a countable family of
atomless Radon probability spaces, and $\tilde\lambda$ the product Radon
probability measure on $X=\prod_{i\in I}X_i$.
If $W\subseteq X$ and $\tilde\lambda W>0$, there is a family
$\familyiI{K_i}$ such that $K_i\subseteq X_i$ is a non-scattered compact
set for each $i\in I$ and $\prod_{i\in I}K_i\subseteq W$.

\proof{{\bf (a)} To begin with, let us suppose that $I=\Bbb N$.   For
each $n\in\Bbb N$, set $Y_n=\prod_{i\ge n}X_i$ and let $\tilde\lambda_n$
be the product Radon probability measure on $Y_n$, so that
$\tilde\lambda_0=\tilde\lambda$ and $\tilde\lambda_n$ can be identified
with the product of $\mu_n$ and $\tilde\lambda_{n+1}$ (417J).   Using
498B repeatedly, we can find non-scattered compact sets
$K_n\subseteq X_n$ and closed non-negligible sets $W_n\subseteq Y_n$
such that $W_0\subseteq W$ and $K_n\times W_{n+1}\subseteq W_n$ for
every $n$.   In this case,
$\prod_{i<n}K_i\times W_n\subseteq W_0$ for every $n$.   If
$x\in\prod_{i\in\Bbb N}K_i$, then there is for each $n\in\Bbb N$ an
$x_n\in\prod_{i<n}K_i\times W_n$ such that $x_n\restr n=x\restr n$, just
because $W_n$ is not empty.   But now every $x_n$ belongs to $W_0$ and

\Centerline{$x=\lim_{n\to\infty}x_n\in W_0\subseteq W$.}

\noindent As $x$ is arbitrary, $\prod_{i\in\Bbb N}K_i\subseteq W$.

\medskip

{\bf (b)} For the general case, we may suppose that $I\subseteq\Bbb N$.
For $i\in\Bbb N\setminus I$, take $(X_i,\frak T_i,\Sigma_i,\mu_i)$ to be
$[0,1]$ with Lebesgue measure.   Set
$\tilde W=\{x:x\in\prod_{i\in\Bbb N}X_i$, $x\restr I\in W\}$.   By (a),
there are non-scattered compact sets $K_i\subseteq X_i$ such that
$\prod_{i\in\Bbb N}K_i\subseteq\tilde W$, in which case
$\prod_{i\in I}K_i\subseteq W$, as required.
}%end of proof of 498C

\exercises{\leader{498X}{Basic exercises (a)}
%\spheader 498Xa
Let $(X,\frak T,\Sigma,\mu)$ be a Radon measure space,
$(Y,\frak S,\Tau,\nu)$ an effectively locally finite $\tau$-additive
topological measure space, and $\tilde\lambda$ the $\tau$-additive
product measure on
$X\times Y$.   Show that if $W\subseteq X\times Y$ and
$\tilde\lambda W>0$ there are a
compact set $K\subseteq X$ and a set $F\subseteq Y$ of positive measure
such that $K\times F\subseteq W$ and $K$ is either non-scattered or
non-negligible.

\spheader 498Xb
Let $(X,\frak T,\Sigma,\mu)$ be an atomless Radon measure space,
$(Y,\Tau,\nu)$ any measure space, and $\lambda$ the c.l.d.\ product
measure on $X\times Y$.   Show that if $W\subseteq X\times Y$ and
$\lambda W>0$ there are a non-scattered
compact set $K\subseteq X$ and a set $F\subseteq Y$ of positive measure
such that $K\times F\subseteq W$.   \Hint{reduce to the case in which
$\nu$ is totally finite and $\Tau$ is countably generated, so that the
completion of $\nu$ is a quasi-Radon measure for an appropriate
second-countable topology.}

\spheader 498Xc Let $\familyiI{(X_i,\frak T_i,\Sigma_i,\mu_i)}$ be any
family of atomless Radon probability spaces, and $\lambda$ the ordinary
product measure on $X=\prod_{i\in I}X_i$.   Show that if $W\subseteq X$
and $\lambda W>0$ then there are non-scattered compact sets
$K_i\subseteq X_i$ for $i\in I$ such that
$\prod_{i\in I}K_i\subseteq W$.
%498C

\spheader 498Xd Let
$\familyiI{(X_i,\frak T_i,\Sigma_i,\mu_i)}$ be a countable family of
atomless Radon probability spaces, and $W\subseteq\prod_{i\in I}X_i$ a
set with positive measure for the Radon product of $\familyiI{\mu_i}$.
Show that there are atomless Radon probability measures $\nu_i$ on $X_i$
such that $W$ is conegligible for the Radon product of
$\familyiI{\nu_i}$.   \Hint{439Xh(vii).}

\leader{498Y}{Further exercises (a)}
%\spheader 498Ya
Let $\familyiI{(X_i,\frak T_i,\Sigma_i,\mu_i)}$ be a family of atomless
perfect probability spaces, and $\lambda$ the ordinary product measure
on $X=\prod_{i\in I}X_i$.   Show that if $W\subseteq X$ and
$\lambda W>0$ then there are sets
$K_i\subseteq X_i$ for $i\in I$, all of cardinal $\frak c$, such that
$\prod_{i\in I}K_i\subseteq W$.
%498C
}%end of exercises

\endnotes{
\Notesheader{498} I have already noted (325Yd) that a set $W$ of
positive measure in a product space need not include the product of two
sets of positive measure;  this fact is also the basis of 419E.   Here,
however, we see that if one of the factors is a Radon measure space then
$W$ does include the product of a
non-trivial compact set and a set of positive measure.   There are many
possible variations on the result, corresponding to different product
measures (498B, 498Xb) and different notions of `non-trivial' (498Xa,
498Ya).
The most important of the latter seems to be the idea of a
`non-scattered' compact set $K$;  this is a quick way of saying that
$[0,1]$ is a continuous image of $K$, which is a little stronger than
saying that $\#(K)\ge\frak c$,
and arises naturally from the proof of 498B.
}%end of notes

\discrpage

