\frfilename{mt41Iconc.tex}
\versiondate{7.10.13}
\copyrightdate{2013}
     
\def\chaptername{Topologies and measures I}
\def\sectionname{Concordance}
\let\headlinesectionname=\sectionname
     
\noindent{\bf Concordance to chapters 41-45}

\medskip
     
I list here the section and paragraph numbers which have 
(to my knowledge) appeared in print in references to these chapters, and 
which have since been changed.
     
\medskip
     
\leader{4{}11I}{Completion regular measures} The definition of `completion regular' 
measure, referred to in {\smc Fremlin 00}, has been moved to 411J.
     
\leader{4{}14N}{Density topologies} The note on `density topologies', 
referred to % 2{}23Yb 
in the 2001 edition of Volume 2, has been moved to 414P.

\leader{4{}15Yd}{Sorgenfrey line} This exercise, referred to in the 2001 
edition of Volume 2, %in 2{}61Yj 
has been moved to 415Ye.

\leader{4{}16M}{Henry's theorem}, mentioned in {\smc K\"onig 04}, 
%following 2.5,
has been re-named 416N.
%416N

\leader{4{}16P}{The algebra of open-and-closed subsets} This paragraph, 
referred to in the 2002 edition of Volume 3, %notes to 3{}62
has been moved to 416Q.

\leader{4{}16T}{Kakutani's theorem} The description of the usual measure 
on $\{0,1\}^{\kappa}$, 
referred to in {\smc Fremlin \& Plebanek 03}, 
has been moved to 416U.

\leader{4{}17E}{$\tau$-additive product measures} The reference in
{\smc Fremlin 00} to Kakutani's theorem that the product measure on
$\{0,1\}^I$ is completion regular should be directed to 415E or 416U 
rather than 4{}17E.
     
\leader{4{}19H} The example of a measure inner regular with respect
to the Borel sets but with no extension to a topological measure, mentioned
in {\smc K\"onig p09}, is now 419J.

\leader{4{}19J}{Partitions into sets of full outer measure}
Lemma 4{}19J, mentioned in the 2004 printing of Volume 1, %134Yr
has been moved to 419I.

\leader{4{}32I}{Capacitability} Definition 4{}32I, 
referred to in the 2008 edition of Volume 5, is now 432J.

\leader{4{}34S-4{}34T}{Vague topologies} The material on vague topologies,
referred to in the 2001 edition of Volume 2, has been moved to \S437.

\leader{4{}39H}{$\tau$-smooth functionals} The example of a $\tau$-smooth
functional which is not representable as an integral, referred to in
{\smc Bogachev 07}, is now 439I.

\leader{4{}39J}{A non-Radon space} The example of a first-countable compact
Hausdorff space which is not Radon,
referred to in {\smc Bogachev 07}, is now 
439K.

\leader{4{}39N}{Baire measure} The example of a Baire probability measure
with no extension to a Borel measure, 
referred to in {\smc Bogachev 07}, is now 439M.

\leader{4{}41A}{Shift actions} The definition of shift actions in 4{}41Ac
of the 2003 and 2006 printings, called on in the 2008 edition of Volume 5,
has been moved to 4A5Cc.
     
\leader{4{}44Xn}{Orthonormal bases} The sketch of a construction of an
orthonormal basis in $L^2$ consisting of equivalence classes of continuous
functions, referred to in {\smc Bogachev 07}, is now 444Ym.

\leader{\S4{}45}{Convolutions} The material mentioned in
the notes to \S257 in the 2001 edition of Volume 2 has been moved to
\S444.   The particular result referred to as `4{}45K' is now 444R.
     
\leader{4{}45Xq}{} The exercise 4{}45Xq, referred to in the
2002 edition of Volume 3, %in 3{}87H 
has been moved to 445Xp.
     
\leader{4{}49I, 4{}49J}{Tarski's theorem} The proof of Tarski's theorem,
referred to in the 2002 and 2004 printings of Volume 3, is now in 449L.     

\leader{45{}2I}{} In {\smc Fremlin 00} I quote Pachl's result that if 
$(X,\Sigma,\mu)$ is countably compact,
$(Y,\Tau,\nu)$ is strictly localizable and $f:X\to Y$ is \imp, then 
$\nu$ is countably compact;  this is now in 452R.
     
\leader{4{}55D}{} The material on Brownian motion in \S4{}55, 
mentioned in {\smc K\"onig 04} and {\smc K\"onig 06},
has been moved to \S477.
% Brownian motion as a Radon measure.  
%\S454 also mentioned

\bigskip

\noindent{\bf References}

\medskip

Bogachev V.I.\ [07] {\it Measure theory.}  Springer, 2007.

Fremlin D.H.\ [00] `Weakly $\alpha$-favourable measure spaces', 
Fundamenta Math.\ 165 (2000) 67-94.
   %n99821.tex
     
Fremlin D.H.\ \& Plebanek G.\ [03] `Large families of mutually singular 
Radon measures', Bull.\ Acad.\ Pol.\ Sci.\ (Math.) 51 (2003) 169-174.

K\"onig H.\ [04] `Projective limits via inner premeasures and the true 
Wiener measure', Mediterranean J.\ Math.\ 1 (2004) 3-42.

K\"onig H.\ [06] `Stochastic processes in terms of inner premeasures', 
Note di Matematica 25 (2006) 1-30.

K\"onig H.\ [p09] `Measure and integration:  the basic extension theorems',
2009 
({\tt www.math.uni-sb.de{\bsp}service{\bsp}preprints/preprint230.pdf}).


\frnewpage
